---
title: "Assignment 3"
author: "Warren Jodjana"
date: "2023-10-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Introduction

Unfortunately, Homicides remain a large issue at hand in large cities all over the world. In this lab, I'm going to be investigating homicides by tracts in two large cities, Los Angeles and New York. In these two cities, incidents of homicides are often not evenly distributed but are clustered in specific areas due to specific socio-economic, environmental and cultural factors. Understanding these spatial distribution can directly influence public safety initiatives and can help for resource allocation for law enforcement and community-based intervention. Both New York and Los Angeles have unique urban histories, patterns of urban development and socio-cultural changes.By analyzing the distribution of homicides in specific tracts, I aim to investigate distinct spatial patterns that might provide information on the underlying causes and potential strategies for mitigation.

## Methods

My analysis started with obtaining and cleaning spatial data for both LA and NYC. The crime data comes from [Los Angeles Open Data Portal](https://data.lacity.org/Public-Safety/Crime-Data-from-2010-to-2019/63jg-8b9z) and [New York Open Data Portal](https://data.cityofnewyork.us/Public-Safety/NYC-crime/qb7u-rbmr/data). The tract data of Los Angeles comes from [City of Los Angeles Hub](https://geohub.lacity.org/search?collection=Dataset&q=Census%20Tracts) and the tract data of New York comes from [NYC Open Data](https://data.cityofnewyork.us/City-Government/2010-Census-Tracts/fxpq-c8ku). In the lab, all crime data is taken from 2015. Once preprocessed, a point-in-polygon (PIP) operation was employed, spatially joining the locations of individual homicide incidents with their corresponding tracts. Challenges encountered included missing data and differing coordinate reference systems (CRS) between the datasets. 


## Data Standardization

Before the spatial analysis could take place, it was necessary to ensure that the datasets were clean and standardized. Firstly, the datasets for NY and LA were filtered to only include homicides from the year 2015. For Los Angeles, it was to filter out "CRIMINAL HOMICIDE" from the dataset. While for New York, it was to filter out both "HOMICIDE-NEGLIGENT-VEHICLE" and "MURDER & NON-NEGL. MANSLAUGHTER", since their both considered homicides. Next, I handled any rows with missing or invalid (NA) latitude and longitude values and removed them. This is to ensure that all remaining data points can be mapped accurately. Then, I subset all the relevant columns including the unique crime identifier and the coordinates to use for analysis. Finally, I ensured that the latitude and longitude values were converted to numeric. Non-numeric coordinates can cause errors in mapping and spatial analysis.

## Spatial Data Operations

The operation used here for spatial data analysis is the point-in-polygon (PIP) method. The PIP method identifies which tracts (polygons) each homicide (point) belongs to. This allows for aggregation of crimes at tract level, effectively associating each tract with a count of homicides. Firstly, the datasets were converted from tabular (CSV) form to spatial point datasets. Then, to make sure that the points and polygons align properly, I transformed both datasets to the same CRS (4326). Then, perform the PIP operation by using spatial join, linking each point to the polygon area it resides. Next, aggregate each homicides by each tract to show the number of homocides within each tract were tallied. Finally, merge all data to showcase a visualization or map.




```{r load-libraries, message=FALSE, include=FALSE}
library(sf)
library(tmap)
library(leaflet)
library(data.table)
library(tidyverse)
```



## Los Angeles Maps


```{r load-data, include=FALSE}
LAcrime <- fread("LA-data/LAPD2015_Violent.csv")

head(LAcrime)
```


```{r unique-code, include=FALSE}
unique(LAcrime$`Crime Code Description`)
```


```{r subset-data, include=FALSE}
LAcrime.df <- as.data.frame(LAcrime)
s1 <- subset(LAcrime.df, LAcrime.df$`Crime Code Description`=="CRIMINAL HOMICIDE") # base r subset
LAcrime.hom <- rbind(s1)

head(s1)
```


```{r save-data, include=FALSE}
write.csv(LAcrime.hom,"LAcrime_hom.csv")
```


```{r xy-locations, include=FALSE}
glimpse(LAcrime.hom[,c("longitude","latitude")])
```


```{r long-lat-str, include=FALSE}
str(LAcrime.hom[,c("longitude", "latitude")])
LAcrime.hom$latitude <- as.numeric(LAcrime.hom$latitude)
LAcrime.hom$longitude <- as.numeric(LAcrime.hom$longitude)
```


```{r missing-data, include=FALSE}
LAcrime.hom.na <- subset(LAcrime.hom, is.na(LAcrime.hom[,c("longitude", "latitude")]))

glimpse(LAcrime.hom.na) # observations
```


```{r missing-values, include=FALSE}
LAcrime.hom2 <- na.omit(LAcrime.hom[,c("DR Number","longitude", "latitude")])
str(LAcrime.hom2)
```


```{r convert-inspect, include=FALSE}
LAcrime.pts <- st_as_sf(LAcrime.hom2, coords = c("longitude","latitude"), crs = 4326)
plot(LAcrime.pts)
```



```{r load-inspect, include=FALSE}
LAcrimes <- LAcrime.pts
LAtracts <- st_read("LA-data/LAC_Shape.shp")

```

### Los Angeles Homicides per Tract

```{r points-polygons, include=TRUE, warning=FALSE, echo=FALSE, message=FALSE}
tmap_mode("plot")
## 1st layer (gets plotted first)
LAptspolygons <- tm_shape(LAtracts) + tm_borders(alpha = 0.4) + 
  
  ## 2nd layer (overlay)
  tm_shape(LAcrime.pts) + tm_dots(size = 0.1, col="red") +
  
  ## add title
  tm_layout(title="Los Angeles Points and Polygons", 
            title.size = 0.65, 
            title.position=c("left", "bottom"), 
            title.fontface = "bold") +
  
  # add scalebar
  tm_scale_bar(breaks = c(0, 5, 10),  
               text.size = 0.8,
               width = 1,             # Adjusts the width of bar segments
               position = c("left", "bottom"),
               lwd = 1.5,             # Adjust line width for visibility
               color.dark = "red",   # Change dark color
               color.light = "lightcoral")  # Change light color

LAptspolygons
```


```{r transform-crs, include=FALSE}
CRS.new <- st_crs(LAtracts)
LAcrimes <- st_transform(LAcrimes, CRS.new)
```


```{r spatial-join, include=FALSE}
crime_in_tract <- st_join(LAcrimes, LAtracts, join = st_within)
glimpse(crime_in_tract)
```


```{r crimes-per-tract, include=FALSE}
crime_tract_count <- as.data.frame(table(crime_in_tract$TRACTCE10))
glimpse(crime_tract_count)
```


```{r rename-columns, include=FALSE}
names(crime_tract_count) <- c("TRACTCE10","CrimeCt")
glimpse(crime_tract_count)

```


```{r merge-data, include=FALSE}
LAtracts_new <- merge(LAtracts, crime_tract_count, by="TRACTCE10")
glimpse(LAtracts_new)

```



### Visualize LA Homicides in 2015

```{r visualize-output, include=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
tmap_mode("view")
tm_shape(LAtracts_new) + tm_fill("CrimeCt", n=4, pal = "BuPu", title="LA Homicides in 2015")
```


## New York City



```{r load-dataNY, include=FALSE}
NYcrime <- fread("NYC-data/NYPD2015_Violent.csv")

head(NYcrime)
```


```{r unique-codeNY, include=FALSE}
unique(NYcrime$`OFNS_DESC`)
```


```{r subsetNY, include=FALSE}
NYcrime.hom <- NYcrime %>%
  filter(`OFNS_DESC` == "HOMICIDE-NEGLIGENT-VEHICLE" | `OFNS_DESC` == "MURDER & NON-NEGL. MANSLAUGHTER")

head(NYcrime.hom)
```


```{r save-dataNY, include=FALSE}
write.csv(NYcrime.hom, "NYcrime_hom.csv")
```


```{r xy-locationsNY, include=FALSE}
glimpse(NYcrime.hom[,c("Longitude","Latitude")])
```


```{r long-latNY, include=FALSE}
str(NYcrime.hom[,c("Longitude", "Latitude")])
NYcrime.hom$Latitude <- as.numeric(NYcrime.hom$Latitude)
NYcrime.hom$Longitude <- as.numeric(NYcrime.hom$Longitude)
```


```{r missing-valuesNY, include=FALSE}
NYcrime.hom2 <- na.omit(NYcrime.hom[,c("CMPLNT_NUM","Longitude", "Latitude")])
str(NYcrime.hom2)
```


```{r convert-inspectNY, include=FALSE}
NYcrime.pts <- st_as_sf(NYcrime.hom2, coords = c("Longitude","Latitude"), crs = 4326)
plot(NYcrime.pts)
```


```{r load-inspectNY, include=FALSE}
NYcrimes <- NYcrime.pts
NYtracts <- st_read("NYC-data/NYC_Shape.shp")

```

### New York City Homicides per Tract
```{r points-polygonsNY, include=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
tmap_mode("plot")
NYptspolygons <- tm_shape(NYtracts) + tm_borders(alpha = 0.4) + 
  
  ## 2nd layer (overlay)
  tm_shape(NYcrime.pts) + tm_dots(size = 0.1, col="blue") +
  
  ## add title
  tm_layout(title="New York City Points and Polygons", 
            title.size = 0.75, 
            title.position=c("left", "top"), 
            title.fontface = "bold") +
  
  # add scalebar
tm_scale_bar(breaks = c(0, 5, 10),  
             width = 1,           # Adjusts the width of bar segments
             position = c("left", "top"),  
             text.size = 0.8,
             lwd = 1.5,            # Adjust line width for visibility
             color.dark = "blue",   # Change dark color
             color.light = "lightblue")  # Change light color

NYptspolygons
```


```{r transform-crsNY, include=FALSE}
CRS.new <- st_crs(NYtracts)
NYcrimes <- st_transform(NYcrimes, CRS.new)
```


```{r spatial-joinNY, include=FALSE}
crime_in_tractNY <- st_join(NYcrimes, NYtracts, join = st_within)
glimpse(crime_in_tractNY)
```



```{r crimes-per-tractNY, include=FALSE}
crime_tract_countNY <- as.data.frame(table(crime_in_tractNY$tractce))
glimpse(crime_tract_countNY)
```


```{r rename-columnsNY, include=FALSE}
names(crime_tract_countNY) <- c("tractce","CrimeCt")
glimpse(crime_tract_countNY)

```


```{r merge-dataNY, include=FALSE}
NYtracts_new <- merge(NYtracts, crime_tract_countNY, by="tractce")
glimpse(NYtracts_new)

```


### Visualize NYC Homicides in 2015
```{r visualize-outputNY, include=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
tmap_mode("view")
tm_shape(NYtracts_new) + tm_fill("CrimeCt", n=4, pal = "BuPu", title="NYC Homicides in 2015")
```

## Discussion

In New York City, the spatial distribution of homicide per tract is much more concentrated especially towards its more central and upper regions which represents Manhattan and its surrounding boroughs. This concentration is likely affected by high population density, socioeconomic conditions and nature of urban environments. Furthermore, Manhattan is considered to be a large CBD (Central Buisness District). This area is a magnet for commerce, entertainment, and residential luxury. While it's home to many wealthy individuals, it also attracts a daily influx of workers, tourists, and visitors from all over the world. In addition, the efficient public transport systems, which brings different groups of people into close proximity, might also increase these patterns. There is also a large concentration of homicides in the Bronx area, which might be tied to economic challenges, leading to poverty, unemployment, and limited access to essential services.The borough's unique urban environment, with a mix of residential housing projects, commercial corridors, and recreational areas, also plays a role in the increase of homicides.

On the other hand, the spatial distribution of homicide per tract is more spread out in Los Angeles, where it's slightly more concentrated in the southern parts of the city. In contrast, central Los Angeles portrays a moderate distribution, indicating a more balanced homicide rate in the city. The sprawling nature of Los Angeles, with its extensive highway system and car-centric culture, means that people often traverse large distances, potentially leading to territorial disputes or gang-related activities. Furthermore, the city's approach to urban planning, with its sprawling neighborhoods, large commercial districts, and pockets of densely populated areas, can influence crime patterns. The connectivity provided by its advanced road network means that there's increased mobility, but it can also lead to isolated incidents in various parts of the city, rather than concentrated hotspots as seen in more compact urban settings.

For future research, incorporating data related to socio-economic status, employment rates, education levels, and community engagement programs might provide more insights into the patterns. Additionally, examining the role of law enforcement presence, community policing, and surveillance in these areas could provide more information on the effectiveness of crime prevention strategies.

















